---
title: "CO2 and GDP"
author: "NM"
date: "08-02-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("ggplot2")
library("ggfortify")
library("forecast")
library("knitr")
##
library('sjstats') #version 0.17.4
##
library('qpcR')
library("gvlma")
library('mctest')
library("lmtest")
library("car") # D-W test
```


## The Problem

Put some economics theory here

## The data

Describe the data

```{r}
rm(list = ls())
### Read all the data
##'NY.GDP.MKTP.KD',    = GDP constant US$ 2015
##'SP.POP.TOTL',       = Population total
##'EN.ATM.CO2E.PC',    = CO2 emission in tons per/capita
##"EG.ELC.RNWX.ZS",    =  Electricity from renovable sources % of total
##'NY.GDP.PCAP.CD',    = GDP per capita (current US$) 
##"NY.GDP.PCAP.PP.CD", = GDP per capita, PPP (current international $) 
##'EG.ELC.COAL.ZS',    = Electricity produced from coal % of total
##'AG.PRD.LVSK.XD',    = Livstock production index (2014--2016=100)
##"AG.CON.FERT.ZS",    = Fertilizer consumption (kg/hectare of arable land)
##"EG.USE.PCAP.KG.OE"  = Energy use (kg of oil equivalent per capita)
wb <- read.csv("https://raw.githubusercontent.com/hrpunio/Erasmus_2021EE/main/WBdata.csv", sep = ';',  header=T, na.string="NA")
### Clean the data
##unwated <- c('OSS', 'LTE', 'AFW', 'LCN', 'LCD')
wb.unwanted <- read.csv("https://raw.githubusercontent.com/hrpunio/Erasmus_2021EE/main/wb_groups_list.csv", 
    sep = ';',  header=T, na.string="NA")
```

More transformations

```{r}
## Transform to _wider_ format
## change colum names to something less complicated
## Crossectional data: year == 2015
## Remove small countries: year >= 1990
wbl <- wb %>% filter (! code %in% wb.unwanted$code ) %>%
  pivot_wider( names_from = indicatorcode, values_from = value) %>%
  dplyr::select (countryname, code, year, 
          co2=EN.ATM.CO2E.PC, 
          gdp_per_cap=NY.GDP.PCAP.PP.CD, 
          e4coal=EG.ELC.COAL.ZS,
          ren_energy = EG.ELC.RNWX.ZS,
          fcons=AG.CON.FERT.ZS, 
          euse=EG.USE.PCAP.KG.OE, 
          livestock = AG.PRD.LVSK.XD,
          gdpt=NY.GDP.MKTP.KD,  
          pop=SP.POP.TOTL) %>%
  dplyr::filter (year == 2015) %>%
  dplyr::filter(pop > 1000000) %>% 
  mutate(log_gdp_per_cap = log(gdp_per_cap),
                   log_co2 = log(co2),
                   gdp_per_cap_squared = gdp_per_cap^2,
                   log_gdp_per_cap_squared = log(gdp_per_cap)^2)

## check sample size:
nrow(wbl)
```

Scatter-plot matrix

```{r}
## scatter-plot matrix for co2, gdp, fcons, euse
wbl %>% dplyr::select(co2, gdp_per_cap, fcons, euse, e4coal) %>% pairs()
```

Correlation matrix for

```{r}
## NOTE: na.omit function omits rows with NA values
wb.corr <- wbl %>% dplyr::select (co2, gdp_per_cap, fcons, euse, e4coal) %>% 
  na.omit() %>% 
  cor()

wb.corr
##
```

## Model 0

$CO2 = a + b GDP$

lin-lin model

```{r}
lm0 <- lm(co2 ~ gdp_per_cap, data=wbl)
summary(lm0)
```

## Model 1

$\ln CO2 = a + b \ln GDP$

log-log model

```{r}
lm1 <- lm(log_co2 ~ log_gdp_per_cap, data=wbl)
summary(lm1)
```

## Model 2

$\ln CO2 = a + b \ln GDP + c \ln^2 GDP $

the environmental Kuznets curve with logs
as the theory of the Kuznets curves suggests, the sign of the coefficient before log_gdp_per_cap is positive and log_gdp_per_cap_squared is negative, resulting in the U-shape
The model is great in terms of R2adj and p-values of the coefficients  

```{r}
lm2 <- lm(log_co2 ~ log_gdp_per_cap + log_gdp_per_cap_squared, data=wbl)
summary(lm2)
```


### chart residuals vs fitted

```{r}
plot(lm2, which = 1) 
```

### Check for heteroscedasticity

Breusch-Pagan Test For Homoscedasticity

```{r}
#high p = no heteroscedasticity
bptest(lm2)
```

### Check for autocorrelation of error term

```{r}
## Check autocorrelation of error term
## Durbin Watson Test for Autocorrelation
## high p = errors not autocorrelated
durbinWatsonTest(lm2)
```


### Check for normality of residuals

```{r}
## visually looking at qq-plot
plot(lm2, which = 2) 
qqPlot(lm2)
shapiro.test(lm2$residuals)
```

## Check multicollinearity
## for models with 

```{r}
omcdiag(lm2)

vif(lm2)
```
## lm2 is highly collinear but it is OK

## Perhaps some other models ...

## Model 3

$\ CO2 = a + b GDP + c GDP^2 $

the classical environmental Kuznets curve 
the sings of the coefficients are concurring with the theory, but the values are very hard to work with - the previous model is clearly better (both by R2adj and p-values of the coefficients)

```{r}
lm3 <- lm(co2 ~ gdp_per_cap + gdp_per_cap_squared, data=wbl)
summary(lm3)
```

## Model 4

after Cederborg, J., & Snöbohm, S. (2016). Is there a relationship between economic growth and carbon dioxide emissions?.

```{r}
lm4 <- lm(co2 ~ gdp_per_cap + gdp_per_cap_squared + ren_energy + e4coal + livestock , data=wbl)
summary(lm4)
```

## Model 5

Another, less theory-driven approach - let the step function find the best model using AIC

```{r}
lm_no_log_full <- lm(co2 ~ gdp_per_cap + gdp_per_cap_squared + ren_energy + e4coal + livestock + fcons + euse + log(gdpt) + log(pop), data=wbl)
lm5 <- stats::step(lm_no_log_full, direction = "backward")
summary(lm5)
```
## Model 6

The same as the previous model, but with logs

```{r}
lm_log_full <- lm(log_co2 ~ log_gdp_per_cap + log_gdp_per_cap_squared + ren_energy + e4coal + livestock, data=wbl)

lm6 <- stats::step(lm_log_full, direction = "backward")
summary(lm6)
```

## For each model perform diagnostics

```{r}
sjstats::check_assumptions(lm2)
```
```{r}
sjstats::check_assumptions(lm3)
```

```{r}
sjstats::check_assumptions(lm4)
```

```{r}
sjstats::check_assumptions(lm5)
```

```{r}
sjstats::check_assumptions(lm6)
```
## Select between nested models with anova
## (significance of difference of variance explainded)

The main problems with using anova function are the following:

1) different defendant variables 

2) because some columns contain NA's, the models are trained 
on different number of rows  => cannot be compared directly

```{r}
anova(lm0, lm3, test="F")
```

```{r}
anova(lm1, lm2, test="F")
```
Concluding remarks: 
However, we can use another approach - rmse for the models with the same dependant variables: 

```{r}
qpcR::RMSE(lm1)
qpcR::RMSE(lm2)
qpcR::RMSE(lm6)
```
```{r}
qpcR::RMSE(lm0)
qpcR::RMSE(lm3)
qpcR::RMSE(lm4)
qpcR::RMSE(lm5)
```

Judging by the combination of the following metrics: 
1) R2adjusted
2) statistical significance of the coefficients
3) RMSE

we can conclude that the model lm6: 
log_co2 ~ log_gdp_per_cap + log_gdp_per_cap_squared + ren_energy + e4coal
is the best model for explaining co2 emissions

## Repeat for other sample

```{r}
wb.highmid <- read.csv('https://raw.githubusercontent.com/hrpunio/Erasmus_2021EE/main/wb_groups.csv', sep = ';',  header=T, na.string="NA") %>%
  ## about 30 countries:
  ##filter ( GroupName == 'Low income')
  ##filter ( GroupName == 'High income')
  dplyr::filter ( GroupName == 'High income' | GroupName == 'Middle income')

nrow(wb.highmid)

wbl_2 <- wb %>% dplyr::filter (code %in% wb.highmid$code ) %>%
  pivot_wider( names_from = indicatorcode, values_from = value) %>%
  dplyr::select (countryname, code, year, 
          co2=EN.ATM.CO2E.PC, 
          gdp_per_cap=NY.GDP.PCAP.PP.CD, 
          e4coal=EG.ELC.COAL.ZS,
          ren_energy = EG.ELC.RNWX.ZS,
          fcons=AG.CON.FERT.ZS, 
          euse=EG.USE.PCAP.KG.OE, 
          livestock = AG.PRD.LVSK.XD,
          gdpt=NY.GDP.MKTP.KD,  
          pop=SP.POP.TOTL) %>%
  mutate(log_gdp_per_cap = log(gdp_per_cap),
                   log_co2 = log(co2),
                   log_pop = log(pop),
                   log_gdp = log(gdpt),
                   gdp_per_cap_squared = gdp_per_cap^2,
                   log_gdp_per_cap_squared = log(gdp_per_cap)^2) %>%
  dplyr::filter (year == 2015)

nrow(wbl)
```

```{r}
lm_1_new  <-  lm(log_co2 ~ log_gdp_per_cap + log_gdp_per_cap_squared, data=wbl_2 ) 
lm_1_new %>% summary()
```
```{r}
lm_2_new  <-  lm(log_co2 ~ log_gdp_per_cap + log_gdp_per_cap_squared + 
    ren_energy + e4coal, data=wbl_2 ) 
lm_2_new %>% summary()
```
```{r}
sjstats::check_assumptions(lm_1_new)
sjstats::check_assumptions(lm_2_new)
```
```{r}
qpcR::RMSE(lm_1_new)
qpcR::RMSE(lm_2_new)
```
Judging by the combination of the following metrics: 

1) R2adjusted

2) statistical significance of the coefficients

3) RMSE

we can conclude that the model lm_2_new: 

log_co2 ~ log_gdp_per_cap + log_gdp_per_cap_squared + ren_energy + e4coal
is the best model for explaining co2 emissions in the High income and Middle income countries
